Optimizing Abstract Abstract Machines

### 2012-11-14 paper/paper.tex
Cleaned up lazy non-determinism

A shortcoming of the AAM approach is that much of the non-determinism
is spurious. For example, in a function application, {\tt (f x y)},
where all variables have been joined with multiple values, each
variable reference transitions to as many new states as there are
values, only to be joined back together in their respective
application positions. We would see transitions of the following
shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

This does not change the concrete semantics of the language. It is an
abstraction of the otherwise several states the original
non-deterministic semantics steps to. We say the abstraction is
\emph{lazy} because it delays dereferencing an address until its
contents are \emph{needed} as values in the semantics. It does not
change the execution order that leads to the values that are stored in
the address.

We introduce a new kind of value, $\saddr{\maddr}$, that represents a
delayed lookup of $\maddr$. This acts to quotient the state space by
the values at the delayed address. The following rules highlight the
changes to the semantics:

@@

\paragraph{Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step. With laziness, the
reference will mean the original binding(s) of the variable or the new
one, because the actual store lookup is delayed one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus might have an observable precision
difference.

\paragraph{Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

>>>>

Tracing the execution of the analysis reveals an immediate shortcoming:
there is a high degree of branching and merging in the exploration.
%
Surveying this branching has no benefit for precision.
%
For example, in a function application, {\tt (f x y)},
where {\tt f}, {\tt x} and {\tt y} each have several values
each argument evaluation induces two-way branching, only to be ultimately joined back together in their respective
application positions. 
%
Transition patterns of this shape litter the state-graph:
%
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
To avoid the spurious forking and joining, we {\it delay} the non-determinism
until and unless it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). 
%
Doing so collapses these forks and joins into a linear sequence of states:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

In case it's unclear, this shift does not change the concrete semantics of the
language to be lazy.  Rather, it abstracts over transitions that the original
non-deterministic semantics steps through. 
%
We say the abstraction is \emph{lazy} because it delays dereferencing an
address until its contents are \emph{needed} as values in the semantics. It
does not change the execution order that leads to the values that are stored in
the address.

We introduce a new kind of value, $\saddr{\maddr}$, that represents a
delayed lookup of $\maddr$.  The following rules highlight the
changes to the semantics:

@@

We have two choices for how to implement lazy non-determinism.

\paragraph{Option 1: Lose precision; simplify implementation}
This semantics introduces a subtle precision difference over the
baseline. Consider a configuration where a reference to a variable and
a binding of a variable will happen in one step. With laziness, the
reference will mean the original binding(s) of the variable or the new
one, because the actual store lookup is delayed one step
(i.e. laziness is administrative). Without laziness, the reference
will fan out to all the bindings of the variable before the new
binding happens and thus might have an observable precision
difference.

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.
